---
title: "R Notebook"
output:
  html_document:
    df_print: paged
    
---

 <font size="7"> Lab 1- Thomas Weil</font> 

<h1> In this Lab I will

<h3>1.Learn to use Rmarkdown


<h3>2.Review R mechanics


<h3>3.Review vector and raster data


&nbsp;





 <font size="4"> Load in Libraries and Check Version</font> 

```{r}
version
library(sf)          # classes and functions for vector data
library(raster) 
library(spData)        # load geographic data
library(spDataLarge)
library(devtools)
library(rasterVis)
devtools::install_github("Nowosad/spDataLarge")

```


&nbsp;

  <font size="4"> Create SP and SF version of World Data frame</font> 
```{r}
library(sp)
world_sp = as(world, Class = "Spatial")
world_sf = st_as_sf(world_sp)
```



&nbsp;

  <font size="4"> Create plot of asia boundaries by making a union of all the different maps of asia</font> 
```{r}
world_asia = world[world$continent == "Asia", ]
asia = st_union(world_asia)
asia
```
&nbsp;



  <font size="4"> Create a plot of the world with circles demarkating population</font>
```{r}
plot(world["continent"], reset = FALSE)
cex = sqrt(world$pop) / 10000
world_cents = st_centroid(world, of_largest = TRUE)
plot(st_geometry(world_cents), add = TRUE, cex = cex)

```


&nbsp;

  <font size="4"> Create and plot new raster</font>
```{r}

new_raster2 = raster(nrows = 6, ncols = 6, res = 0.5, 
                     xmn = -1.5, xmx = 1.5, ymn = -1.5, ymx = 1.5,
                     vals = 1:36)
plot(new_raster2)



```