---
title: "Lab 2"
author: "Thomas Weil GIS 3"
output:
  html_document:
    theme: cosmo
    toc: true
    df_print: paged
  pdf_document: default
---

In this lab we will be constructing a clustered index, to show areas that are similar in terms of several demographic, housing and socio-economic variables. This will allow us to see the degree to wich certain areas in the city are like other areas. 

## Load in Data

Create a data frame of Numerators and Denomenators
```{r}
setwd("/Users/thomasweil/Desktop/YEAR2/ZOOM university/GIS 3/lab 6")
load("census_2011_UK_OA.RData")
Census_2011_Count <- merge(Liverpool,Census_2011_Count_All,by="OA",all.x=TRUE)
head(OAC_Input_Lookup[,])
#(We will now write some code that will calculate all of the numerators:



OAC_Input <- as.data.frame(Census_2011_Count$OA)
colnames(OAC_Input) <- "OA"


#For loop in order to create the correct variables

for (n in 1:nrow(OAC_Input_Lookup)){
      # Get the variables to aggregate for the row specified by n
      select_vars <- OAC_Input_Lookup[n,"England_Wales"]
      
      # Create a list of the variables to select
      select_vars <- unlist(strsplit(paste(select_vars),","))
      
      # Create variable name
      vname <- OAC_Input_Lookup[n,"VariableCode"] 
      
      # Creates a sum of the census variables for each Output Area
      tmp <- data.frame(rowSums(Census_2011_Count[,select_vars, drop=FALSE]))
      colnames(tmp) <- vname
      
      # Append new variable to the OAC_Input object
      OAC_Input <- cbind(OAC_Input,tmp)
      
      # Remove temporary objects
      remove(list = c("vname","tmp"))
} 
OAC_Input$k035 <- NULL


OAC_Input_den <- as.data.frame(Census_2011_Count$OA)
colnames(OAC_Input_den) <- "OA"


den_list <- unique(OAC_Input_Lookup[,"Denominator"])
den_list <- paste(den_list[den_list != ""])
# Select denominators
OAC_Input_den <- Census_2011_Count[,c("OA",den_list)]

OAC_Input <- merge(OAC_Input,OAC_Input_den, by="OA")
```

## More Data Frame Manipulation
Create numerator denominator list

```{r}

K_Var <- OAC_Input_Lookup[OAC_Input_Lookup$Type == "Count",c(1,3)]

head(K_Var)
```

##Run a loop to compare numerator and denomentaor and create percentages

```{r}
# Create an OA list / data frame
OAC_Input_PCT_RATIO <- subset(OAC_Input, select = "OA")
# Loop
for (n in 1:nrow(K_Var)){
  
  num <- paste(K_Var[n,"VariableCode"]) # Get numerator name
  den <- paste(K_Var[n,"Denominator"]) # Get denominator name
  tmp <- data.frame(OAC_Input[,num] / OAC_Input[,den] * 100) # Calculate percentages
  colnames(tmp) <- num
  OAC_Input_PCT_RATIO <- cbind(OAC_Input_PCT_RATIO,tmp) # Append the percentages
  
  # Remove temporary objects
  remove(list = c("tmp","num","den"))
}
```


```{r}
# Create an OA list / data frame
OAC_Input_PCT_RATIO <- subset(OAC_Input, select = "OA")
# Loop
for (n in 1:nrow(K_Var)){
  
  num <- paste(K_Var[n,"VariableCode"]) # Get numerator name
  den <- paste(K_Var[n,"Denominator"]) # Get denominator name
  tmp <- data.frame(OAC_Input[,num] / OAC_Input[,den] * 100) # Calculate percentages
  colnames(tmp) <- num
  OAC_Input_PCT_RATIO <- cbind(OAC_Input_PCT_RATIO,tmp) # Append the percentages
  
  # Remove temporary objects
  remove(list = c("tmp","num","den"))
}
```

## Load in census table, merge to find density
Use merge function

```{r}
#Extract Variable
tmp <- Census_2011_Count[,c("OA","KS101EW0008")]
colnames(tmp) <- c("OA","k007")
#Merge
OAC_Input_PCT_RATIO <- merge(OAC_Input_PCT_RATIO,tmp,by="OA")
```

## Calaculate Standardized Illness Rate


```{r}
# Calculate rates of ill people 15 or less and greater than or equal to 65
ill_16_64 <- rowSums(Census_2011_Count[,c("KS301EW0005","KS301EW0006")]) # Ill people 16-64
ill_total <-   rowSums(Census_2011_Count[,c("KS301EW0002","KS301EW0003")]) # All ill people
ill_L15_G65 <- ill_total - ill_16_64 # Ill people 15 or less and greater than or equal to 65
# Calculate total people 15 or less and greater than or equal to 65
t_pop_16_64 <- rowSums(Census_2011_Count[,c("KS102EW0007","KS102EW0008","KS102EW0009","KS102EW0010","KS102EW0011","KS102EW0012","KS102EW0013")]) # People 16-64
t_pop <- Census_2011_Count$KS101EW0001 # All people
t_pop_L15_G65 <- t_pop - t_pop_16_64 # All people 15 or less and greater than or equal to 65
# Calculate expected rate
ex_ill_16_64 <- t_pop_16_64 * (sum(ill_16_64)/sum(t_pop_16_64)) # Expected ill 16-64
ex_ill_L15_G65 <- t_pop_L15_G65 * (sum(ill_L15_G65)/sum(t_pop_L15_G65)) # Expected ill people 15 or less and greater than or equal to 65
ex_ill <- ex_ill_16_64 + ex_ill_L15_G65 # total expected ill people
# Ratio
SIR <- as.data.frame(ill_total / ex_ill * 100) # ratio between ill people and expected ill people
colnames(SIR) <- "k035"
# Merge data
OAC_Input_PCT_RATIO <- cbind(OAC_Input_PCT_RATIO,SIR)
# Remove unwanted objects
remove(list=c("SIR","ill_16_64","ill_total","ill_L15_G65","t_pop_16_64","t_pop","t_pop_L15_G65","ex_ill_16_64","ex_ill_L15_G65","ex_ill"))
OAC_Input_PCT_RATIO 
```



```{r}
# Calculate inverse hyperbolic sine
OAC_Input_PCT_RATIO_IHS <- log(OAC_Input_PCT_RATIO[,2:61]+sqrt(OAC_Input_PCT_RATIO[,2:61]^2+1))
# Calculate Range
range_01 <- function(x){(x-min(x))/(max(x)-min(x))} # range function
OAC_Input_PCT_RATIO_IHS_01 <- apply(OAC_Input_PCT_RATIO_IHS, 2, range_01) # apply range function to columns
# Add the OA codes back onto the data frame as row names
rownames(OAC_Input_PCT_RATIO_IHS_01) <- OAC_Input_PCT_RATIO$OA
```

## Estimating the optimal number of clusters

Use sum of squares technique to calculate optimal number of clusters

```{r}
library(ggplot2)
# Create a new empty numeric object to store the wss results
wss <- numeric()
# Run k means for 2-12 clusters and store the wss results
for (i in 2:12) wss[i] <- sum(kmeans(OAC_Input_PCT_RATIO_IHS_01, centers=i,nstart=20)$withinss)
# Create a data frame with the results, adding a further column for the cluster number
wss <- data.frame(2:12,wss[-1])
# Plot the results
names(wss) <- c("k","Twss")
ggplot(data=wss, aes(x= k, y=Twss)) + geom_path() + geom_point() + scale_x_continuous(breaks=2:12) + labs(y = "Total within sum of squares")
```

We will chose 7 clusters

## Building the geodemographic

No need to run this, it simple creates optomized clusters. 

```{r eval= FALSE}
cluster_7 <- kmeans(x=OAC_Input_PCT_RATIO_IHS_01, centers=7, iter.max=1000000, nstart=10000)
OAC_Input_PCT_RATIO_IHS_01
```

## Load in the clusters premade from another data set

```{r}
setwd("/Users/thomasweil/Desktop/YEAR2/ZOOM university/GIS 3/lab 6")
# Load cluster object
load("cluster_7.Rdata")
```






## Examine Clusters

```{r}
# Lookup Table
lookup <- data.frame(cluster_7$cluster)
# Add OA codes
lookup$OA <- rownames(lookup)
colnames(lookup) <- c("K_7","OA")
# Recode clusters as letter
lookup$SUPER <- LETTERS[lookup$K_7]
```

We will also look at the distribution of these clusters:

```{r}
table(lookup$K_7)
```

## Map The Clusters 

Use tmap
```{r}
# Load packages
library(rgdal)
library(tmap)
# Import OA boundaries
liverpool_SP <- readOGR("Liverpool_OA_2011.geojson", "OGRGeoJSON")
# Merge lookup

plot(liverpool_SP)
liverpool_SP@data

liverpool_SP <- merge(liverpool_SP, lookup, by.x="oa_code",by.y="OA")

tmap_mode("plot")
tmap_style("beaver")
m <- tm_shape(liverpool_SP, projection=27700) +
    tm_polygons(col="SUPER", border.col = "grey50",   palette="Set2",border.alpha = .3, title="Cluster", showNA=FALSE) +
  tm_layout(legend.outside=TRUE,legend.position = c("left", "bottom"), frame = FALSE, main.title= "Liverpool Cluster Analysis", bg.color="white")+tm_compass(position= c("right", "top"), color.dark="green", color.light="pink", text.color= "red", size=2)+tm_credits("urban_analytics/10_Data_Reduction_Geodemographics/P10_Data_Reduction_Geodmeographics.Rmd
")+tm_basemap("Stamen.Watercolor")

m


#Create leaflet plot

```

This map, showing the areas that belong to each data cluster, shows some clear spatial clustering as well. The downtown area has many areas in the G category, then there seem to be a few rings, descending down to f, e, C, and then the rest. The d tend to be farther out. There is an interesting cluster of the E cluster around the F and and G but out near the water. This is porbbably due to specific enviormental conditions/job oppurtunities specific to those groups of people (fishing?). These are my current take aways from the map. 
